Emittance preservation of an electron bunch in a loaded quasi-linear plasma wakefield

### 2017-08-16 Tex/Paper.tex
Added beam quality figure
Updated beamQuality figure and added one with absolute values. Also added relevant text

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
$\lambda_{p} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$, will drive an
increasingly strong field $E_{z}$ for each bunch \cite{chen:1985}. A tailing witness beam loading the peak accelerating
phase of this field will quickly gain energy from the wakefield. Acceleration of an electron witness beam driven by two
electron drive beams was demonstrated at Brookhaven National Laboratory \cite{muggli:2011}. However, electron drive
bunches will quickly lose energy to the plasma, and the accelerating witness beam will eventually catch up with the
drive beam and the acceleration stop.

@@

The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam as its driver. The first run
of the experiment is set up with a $10\unit{m}$ rubidium plasma cell, with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. The corresponding plasma wavelength is $1.26\unit{mm}$. The plasma
density is matched to the SPS proton beam such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = \lambda_{pe}/2\pi$ is the
plasma wave number. The aim of the first run is to demonstrate self-modulation of the proton beam, and in 2018 to
sample the wake field with a long electron bunch. The study presented here is for run two, which aims to demonstrate
acceleration of a short electron bunch producing at high energy and low emittance and low energy spread.

@@

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations. The proton beam was pre-modulated by a clipped cosine function to the
longitudinal density profile, with a period matching $\lambda_{pe}$. Its transverse profile was kept as a Gaussian with
$\sigma_{r} = k_{pe} = 200\unit{\mu m}$. The length of the simulation beam was limited to $26\cdot\lambda_{pe}$, and the
electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed several parameter scans with
this setup, testing for the the maximum beam charge that could be accelerated with a high energy gain and low energy
spread \cite{adli:2016, berglyd_olsen:2016}.

@@

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam.
This condition is replicated in out single bunch case as the peak density of the drive beam is $0.83\cdot n_{0}$. The
single bunch setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size
$\sigma_{r} = k_{pe} = 200\unit{\mu m}$. The length of the bunch $\sigma_{z} = 40\unit{\mu m}$.

@@

@@

The large number of parameters involved makes the problem difficult to solve in simulations, and difficult to measure in
experiments. For this study we limited ourselves to look at a range of initial emittances, which allows for a range of
transverse beam sizes when with a fixed $\beta$-function matched to the plasma density; and a range of beam lengths with
a maximum at the same point on the wakefield as for our reference case covered above.

@@

These numerical simulations have been made possible through access to the Abel computing cluster in Oslo, Norway. Abel
is maintained by UNINETT Sigma2 AS and financed by the Research Council of Norway, the University of Bergen, the
University of Oslo, the University of Tromsø and the Norwegian University of Science and Technology. Project code for
this work is nn9303k.

>>>>

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
\begin{align}
    \lambda_{p} = 2\pi\frac{c}{\omega_{pe}} \label{EQ:LambdaP}
\end{align}
and
\begin{align}
    \omega_{pe} = \sqrt{\frac{n_{0} e^{2}}{m_{e} \epsilon_{0}}}, \label{EQ:OmegaP}
\end{align}
%$\lambda_{p} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$,
will drive an increasingly strong field $E_{z}$ for each bunch \cite{chen:1985}. A tailing witness beam loading the peak
accelerating phase of this field will quickly gain energy from the wakefield. Acceleration of an electron witness beam
driven by two electron drive beams was demonstrated at Brookhaven National Laboratory \cite{muggli:2011}. However,
electron drive bunches will quickly lose energy to the plasma, and the accelerating witness beam will eventually catch
up with the drive beam and the acceleration stop.

@@

The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam as its driver. The first run
of the experiment is set up with a $10\unit{m}$ rubidium plasma cell, with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. The corresponding plasma wavelength is $1.26\unit{mm}$. The plasma
density is matched to the SPS proton beam such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the
plasma wave number. The aim of the first run is to demonstrate self-modulation of the proton beam, and in 2018 to
sample the wake field with a long electron bunch. The study presented here is for run two, which aims to demonstrate
acceleration of a short electron bunch producing at high energy and low emittance and low energy spread.

@@

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations. The proton beam was pre-modulated by a clipped cosine function to the
longitudinal density profile, with a period matching $\lambda_{pe}$. Its transverse profile was kept as a Gaussian with
$\sigma_{r} = k_{pe}^{-1} = 200\unit{\mu m}$. The length of the simulation beam was limited to $26\cdot\lambda_{pe}$,
and the electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed several parameter
scans with this setup, testing for the the maximum beam charge that could be accelerated with a high energy gain and low
energy spread \cite{adli:2016, berglyd_olsen:2016}.

@@

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam.
This condition is replicated in out single bunch case as the peak density of the drive beam is $0.83\cdot n_{0}$. The
single bunch setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size
$\sigma_{r} = 200\unit{\mu m}$. The length of the bunch $\sigma_{z} = 40\unit{\mu m}$.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQuality}
    \caption{\label{Fig:BeamQ} Ratio of total beam charge with an emittance growth $\Delta\epsilon \leq 5\%$ as a
        function of initial beam charge (blue), and relative energy spread (red), after $4\unit{m}$ of plasma and with
        an initial emittance $\epsilon_{N,0}=2\unit{\mu m}$. These are shown for four different $\sigma_{z}$ from
        $40\unit{\mu m}$ to $100\unit{\mu m}$. The detailed studies presented in beam loading section corrspond to the
        square marked lines at $100\unit{pC}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQualityAbs}
    \caption{\label{Fig:BeamQAbs} Total beam charge with an emittance growth $\Delta\epsilon \leq 5\%$ as a function of
        initial beam charge (blue), and final momentum (red), after $4\unit{m}$ of plasma and with an initial emittance
        $\epsilon_{N,0}=2\unit{\mu m}$.}
\end{figure}

@@

The large number of parameters involved makes the problem difficult to solve in simulations, and difficult to measure in
experiments. Presented in figures \ref{Fig:BeamQ} and \ref{Fig:BeamQAbs} is a parameter scan for a matched beam with
initial normalised emittance $\epsilon_{N,0}=2\unit{\mu m}$ after propagating through $4\unit{m}$ of plasma. We ran the
scan with beam charge from $10\unit{pC}$ to $300\unit{pC}$ and with length $\sigma_{z}$ from $40\unit{\mu m}$ to
$100\unit{\mu m}$. We see from figure \ref{Fig:BeamQ} that both the $40\unit{\mu m}$ and the $60\unit{\mu m}$ beam has a
well defined minimum energy spread with a beam charge $\geq 50\unit{pC}$ and $\approx 100\unit{pC}$ respectively. Lower
beam charges tend to underload the field, while higher beam charges tend to overload. It is also clear that long beams
with respect to the accelerating flank of the $E_{z}$-field, $\approx\lambda_{pe}/4$, will not optimally load the field
along its length thus experiencing a larger spread in energy.

A higher charge beam will generate a non-linear wakefield sooner, which in turns ensures that a smaller portion of the
beam experiences the quasi-linear region with emittance growth. As figure \ref{Fig:BeamQAbs} illustrates, a higher
charge also results in a lower energy gain.

@@

These numerical simulations have been made possible through access to the Abel computing cluster in Oslo, Norway. Abel
is maintained by UNINETT Sigma2 AS and financed by the Research Council of Norway, the University of Bergen, the
University of Oslo, the University of Tromsø and the Norwegian University of Science and Technology. Project code:
nn9303k. Some of the simulations were also run on the student-maintained computing cluster ``Smaug'' at the University
of Oslo, Department of Physics.
